---
title: "Day 4 Activities"
subtitle: "Intro to functions"
output: 
  html_document:
    css: activity_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(kableExtra)
```

## Part 0. Setup

- Create a repo on GitHub named `eds221-day4-activities`
- Clone to create a version-controlled R Project
- Create some subfolder infrastructure (docs, data)
- Add a new R Markdown file, save as `loops_and_functions.Rmd` in `docs`

## Part 1. For loops revisited

### Task 1

Create two sequences, one called `weekdays` that contains days of the week ("Monday", "Tuesday", "Wednesday", etc.) and one called `transects` that contains the series of transect names "Transect A", "Transect B, "Transect C". Write a nested for loop that creates a matrix containing the following: 


```{r, echo = FALSE}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

transects <- c("Transect A", "Transect B", "Transect C")

study_matrix <- matrix(ncol = length(transects), nrow = length(weekdays))

for (i in seq_along(weekdays)) {
  for (j in seq_along(transects)) {
    study <- paste0(weekdays[i], " - ", transects[j])
    study_matrix[i, j] <- study
  }
}

study_matrix %>% 
  kable() %>% 
  kable_styling()
```

## Part 2. Functions introduced

### Task 2

Write a function called `force` that calculates a force (in Newtons), given inputs of mass (in kg) and acceleration (in $\frac{m}{s^2}$ (recall: $F = ma$), and returns a statement "The resulting force is ___ Newtons."

```{r, include = FALSE}
force <- function(mass, acceleration) {
  force_val <- mass * acceleration
  print(paste("The resulting force is", force_val, "Newtons."))
}

force(100, 2.9)
```


### Task 3

Write a function called `name_column` that, given a data frame name and a number, will return the column name associated with that column number. Add an error message if the column number is < 1, or if the number exceeds the number of columns in the data frame. Test the function using the `mtcars` data frame.

```{r, include = FALSE}
name_column <- function(df, col_no) {
  if (col_no < 1 | col_no >= length(df)) {
    stop("Column does not exist in the data frame.")
  }
  names(df[col_no])
}
```

### Task 3

Write a function called `fishmass`, which returns the weight of a fish given two inputs: common name and length. As a reminder, the length:weight relationship for fish is: $W=aL^b$, where where L is total fish length (centimeters), W is the expected fish weight (grams), and a and b are species-dependent parameter values (shown below for several fish from Peyton et al. 2016).

```{r, echo = FALSE}
fish_parms <- tribble(
  ~sci_name, ~common_name, ~a_est, ~b_est,
  "Chanos chanos", "Milkfish", 0.0905, 2.52,
  "Sphyraena barracuda", "Great barracuda", 0.0181, 3.27,
  "Caranx ignobilis", "Giant trevally", 0.0353, 3.05
)

fish_parms %>% 
  kable() %>% 
  kable_styling(full_width = FALSE)
```
Recreate the table above as a data frame stored as `fish_parms`. Then, write a function called `fish_weight` that allows a user to only enter the *common name* (argument `fish_name`) and *total length* (argument `tot_length`) (in centimeters) of a fish, to return the expected fish weight in grams. Test it out for different species and lengths. 

```{r, include = FALSE}

fish_weight <- function(fish_name, tot_length) {
  fish_sub <- filter(fish_parms, common_name == fish_name)
  wt <- fish_sub$a_est * (tot_length ^ fish_sub$b_est)
  return(wt)
}

fish_weight(fish_name = "Milkfish", tot_length = 57)

```

Now, try creating a vector of lengths (e.g. 0 to 100, by increments of 1) and ensuring that your function will calculate the fish weight over a *range* of lengths for the given species (try this for milkfish, storing the output weights as `milkfish_weights`.

```{r, include = FALSE}
# Create the vector of lengths
lengths_vector <- seq(from = 0, to = 100, by = 1)

# Send it to the function: 
milkfish_weights <- fish_weight(fish_name = "Milkfish", tot_length = lengths_vector)
```

### Task 4

Wave power (more accurately wave energy flux) in deep water is approximated by: 

$$P_{deep}=0.5 H^2 T$$
where $P$ is power in $\frac{kW}{m}$ (potential power per wave meter), $H$ is wave height in meters (more specifically, the *significant wave height*), and $T$ is the wave period in seconds. Learn more [here](https://en.wikipedia.org/wiki/Wave_power).

Write a function called `wave_power` that calculates potential ocean wave power given inputs of wave height and period. 

```{r, echo = FALSE}
wave_power <- function(wave_height, wave_period) {
  0.5 * (wave_height ^ 2) * wave_period
}
```

Use your `wave_power` function to approximate wave power for a period of 8 seconds, over a range of wave heights from 0 to 3 meters by increments of 0.2 meters.

```{r, include = FALSE}
wave_heights <- seq(from = 0, to = 3, by = 0.2)

wave_power(wave_height = wave_heights, wave_period = 8)

```

### Task 5

The wave energy equation changes based on ocean depth. Along the coast of Brenville, which has a very sharp shelf as the wave approaches the coast, wave energy is approximated using the deep ocean equation (the one you used above) for depths > 12 meters, and a shallow equation for depths <= 12 meters. The Brenville team estimates shallow wave power by:

$$P_{shallow}=0.81 H^2 T$$

Create a function that requires inputs of water depth, wave height and period, then returns the approximated wave power using the correct equation for the depth entered. It should also include a message (hint: use `message()` just like you would use warning!) that lets the user know if the shallow or deep water equation was used. 

```{r, include = FALSE}

brenville_waves <- function(wave_height, wave_period, depth) {
  
  if (depth > 12) {
    message("Using the deep wave equation.")
  }
  
  if (depth <= 12) {
    message("Using the shallow wave equation.")
  }
  
  if (depth > 12) {
    bren_wave_energy <- 0.5 * (wave_height ^ 2) * wave_period
  } else if (depth <= 12) {
    bren_wave_energy <- 0.81 * (wave_height ^ 2) * wave_period
  }
  return(bren_wave_energy)
}

brenville_waves(wave_height = 5, wave_period = 10, depth = 15)
```

<!---
### Task 6

Bacterial growth is modeled by $$C_t = C_0e^{rt}$$

where $C_t$ is the estimated population size at any time $t$, $C_0$ is the original population (cells / mL), $r$ is the intrinsic growth rate (usually in 1 / hr), and $t$ is time duration in hours. 

Write a function that has arguments for a starting population size `pop_0` and a maximum time duration over which to model growth `t_max` (in hours). Create a function that accepts those arguments, then estimates the population from time = 0 to time = t_max (by increments of 1 hour), for growth rates from 0.10 to 0.50 (by increments of 0.05). 

```{r}


bacteria_fun <- function(pop_0, t_max) {
  
  rate_sequence <- seq(from = 0.10, to = 0.50, by = 0.05)
  time_sequence <- seq(from = 0, to = t_max, by = 1)
  
  for (i in seq_along(rate_sequence)) {
    for (j in seq_along(time_sequence)) {
      pop_t <- pop_0 * exp(rate_sequence[i] * time_sequence[j])
      print(pop_t)
    }
  }
}

my_test <- bacteria_fun(pop_0 = 5000, t_max = 20)
```

--->



### Citations

Peyton, K. A., T. S. Sakihara, L. K. Nishiura, T. T. Shindo, T. E. Shimoda, S. Hau, A. Akiona, and K. Lorance. 2016. “Length–Weight Relationships for Common Juvenile Fishes and Prey Species in Hawaiian Estuaries.” Journal of Applied Ichthyology 32 (3): 499–502. https://doi.org/10.1111/jai.12957. 
