---
title: "Day 2 Activities"
subtitle: "Data types, indexing data, import & plot data"
output: 
  html_document:
    css: activity_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
```

## Part 1. Checking data types

### Setup

- Create a new repo on GitHub for today's activities
- Clone to create a version controlled R Project 
- Create subfolders called `docs`, `data`, and `figs`
- Create a Quarto document, save in the `docs` subfolder as `r_data_types.qmd` 

### Create some data, check the classes, index!

#### Vectors, lists & data frames

In your Quarto document: 

1. Create a vector called `vec_1`containing the following: 

```{r, eval = FALSE}
2, 5, 9, 10, 8, 12, 1, 0
```

```{r, include = FALSE}
vec_1 <- c(2, 5, 9, 10, 8, 12, 1, 0)

vec_1_e5_up <- vec_1[5:7]

vec_1_char <- as.character(vec_1)
vec_1_char

vec_1_logical <- as.logical(vec_1)
vec_1_logical
```

Check the following for that vector: 

- What is the class of the vector? `class()`
- What type of variable does it store? `typeof()`
- Access the 3rd element and store as `vec_1_e3`
- Access the 1st element and store as `vec_1_e1`
- Access the 5th through 7th elements and store as `vec_1_e5to7`
- Reassign `vec_1` as a character using `as.character`, stored as `vec_1_char`. What does the output look like? 

2. Create a vector called `vec_2`

`vec_2` should contained *named* elements, where `town = "Santa Barbara`, `location = "Rincon"`, `swell = "south"`

```{r, include = FALSE}
vec_2 <- c(town = "Santa Barbara", location = "Rincon", swell = "south")

class(vec_2)
typeof(vec_2)

vec_2[2]

vec_2["location"]

```

- Take a look at what you've made
- What is the class of vector elements? `class()`
- What is the length of vec_2?
- Access the 2nd element by name and store as `vec_2_e2`

3. Create a data frame in R, index

Write code to create a data frame called `df_1` that looks like this: 
```{r, echo = FALSE}
df_1 <- data.frame(region = c("A", "B", "A", "D"),
                   species = c("otter", "great white", "sea lion", "gray whale"),
                   count = c(12, 2, 36, 6))

df_1

max_count <- max(df_1$count)
```

- Return the class of the entire data frame
- Return the class of the species column
- Find the maximum value of the `count()` column, store as `max_count`


## Part 2. Wild data

#### Set-up

- Visit the EDI site to learn about Mack Creek salamander & cutthroat trout data you'll be using here: [data package](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-and.4027.14)

- Download the *first CSV* listed (`AS00601.csv`), and take a look at it (outside of R is fine as a first step, e.g. you can open the CSV in Excel)

- Explore the metadata (see View Full Metadata in the [Resources section](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-and.4027.14) of the data website)

- What does each column contain? What are the units of each? What is the study overall about? 

- Create a new Quarto document and save it in your `docs` folder. Attach the `tidyverse`, `here` and `janitor` packages in the setup chunk (you choose the file name)

- Set global options in the YAML so that messages and warnings do NOT show up in the rendered document

- Save the `AS00601.csv` in your `data` folder of your project

#### Read in the data

- Read in the data using `read_csv()` with `here()`, store as `mack_verts`
```{r, include = FALSE}
mack_verts <- read_csv(here("activities","AS00601.csv"))
```

- Look at what you've read in (e.g. with view())

#### A bit of wrangling & exploring

- Update the variable names in mack_verts to lower snake case

```{r, include = FALSE}
mack_verts <- clean_names(mack_verts)
```

- In a new code chunk, practice accessing individual pieces of the data frame (there is no real functionality to this right now, but just to reinforce stuff we learned in our interactive session): 

  - Store the 5th value in column "WEIGHT" as `mc_wt_5`. Check by looking at your data frame to confirm.
  - Store the 8th - 20th value in the "LENGTH1" column as `mc_length_8_20`. Check by looking at your data frame to confirm. 
  - Store everything in column SAMPLEDATE as a vector called `mc_dates`
    
```{r, include = FALSE}
mc_wt_5 <- mack_verts$weight[5]

mc_length_8_20 <- mack_verts$length1[8:20]

mc_dates <- mack_verts$sampledate
```

#### Make a salamander subset

- Create a subset that only contains observations for Pacific Giant Salamanders (species *Dicamptodon tenebrosus*, stored in `species` as `DITE`). Store the subset as `mc_salamanders`. *Hint:* see `dplyr::filter()`!

```{r, include = FALSE}
mc_salamanders <- mack_verts %>% 
  filter(species == "DITE")
```

#### Make a scatterplot of salamander length x weight

- Create a scatterplot of length1 (snout-vent length in millimeters) versus weight (grams) for all salamanders in the subset you created above, `mc_salamanders`. Update axis labels, title, subtitle, and add a caption with the data source. Customize point color and size, possibly opacity, and theme. 

```{r, include = FALSE}
salamander_plot <- ggplot(data = mc_salamanders, aes(x = length1, y = weight)) + 
  geom_point(alpha = 0.5, color = "purple") +
  theme_minimal()

salamander_plot
```

- Export your scatterplot as `salamander_size.png` to your figs folder.

```{r, include = FALSE}
ggsave(here::here("activities", "salamander_size.png"), plot = salamander_plot)
```

#### Make a cutthroat plot

- Similar to above, make a subset called `mc_trout` that only contains observations for cutthroat trout (species "ONCL")
- Create a scatterplot of length1 by weight for all trout in the dataset
- Customize so that the point color depends on *reach* 
- Customize your color scheme (e.g. `scale_color_manual()`)
- Facet your plot by creek reach (`facet_wrap(~...)`)
- Update graph axis labels and title
- Export your graph as `cutthroat_size.png` to the `figs` folder

```{r, include = FALSE}
mc_trout <- mack_verts %>% filter(species == "ONCL")

ggplot(data = mc_trout, aes(x = length1, y = weight)) +
  geom_point(aes(color = reach)) +
  scale_color_manual(values = c("red", "orange", "purple")) +
  facet_wrap(~reach) +
  theme_minimal()

ggsave(here("activities", "cutthroat_size.png"))
```

### Stage, commit, pull, push

- Make sure your changes are safely stored by pushing to GitHub
- Close your project locally
- Reopen your project locally
- Reopen your .qmd files for the activities you did today.
- Render. Does it work? Done. 

## End Day 2 activities
