---
title: "Long-term ice cover duration: North Temperate Lakes LTER"
output: 
  html_document:
    css: activity_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = FALSE)
library(tidyverse)
library(skimr)
library(lubridate)
library(here)
library(kableExtra)
```

In this activity, you'll explore ice cover duration for lakes in the Madison Lake Area (long term data, 1853 - present), then further explore trends in mean air temperature from a second dataset. 

The datasets are here to download: 

- [Madison Lake Area - ice duration data](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-ntl.33.35)
- [Madison, Wisconsin meteorological data](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-ntl.20.32)

## Your task: 

- Download both datasets

- Explore the metadata for each to ensure you know what they contain, variables, units, etc. **NOTE:** How are missing values stored in the ice duration variable?

- Create a new version-controlled R Project 

- In a single R Markdown (or Quarto) document:

  - Read in both files (ice cover duration and meteorological data). Note that the ice cover data has some NA values stored as -999. 
  
  - Create a visualization that compares ice duration across different lakes in the Madison Lake Area. Include both the actual ice duration values *and* some representation of summary statistics for each lake. Careful if you use a boxplot over a jitter or beeswarm - what values might be doubly represented? Can you find an argument that would help you remove those from the boxplot? 
  
  - Find the mean ice cover duration in Madison Lake Area by year (i.e., you'll only be grouping by the year variable here). Understanding check: what are you finding the mean of here? Why are there multiple observations for each year?)
  
  - Create an exploratory visualization of mean ice cover duration by year for the Madison Lake Area. Add a sentence or two below this exploratory graph describing the overall trend(s) you observe. 
  
  - Find the mean air temperature (using the daily average air temperature - adjusted) in Madison Lake Area **by year** using **only observations from winter months (December, January, February)**
  
  - Create an exploratory visualization of mean winter temperatures for the Madison Lake Area. Add a sentence or two below this exploratory graph describing the overall trend(s) you observe.
  
  - Join the mean winter air temperatures to the mean ice cover duration data you found above
  
  - Create an exploratory scatterplot of mean winter temperatures versus mean ice duration. Add a sentence or two below this exploratory graph describing the overall trend(s) you observe. 
  
  - Explore thaw dates since 1970. Create a subset of the ice cover duration data since 1970 (ok to include 1970), then convert the `ice_off` column to a `Date`. Use `lubridate::yday()` to pull the numeric *day of the year* from that Date. Create an exploratory plot of the ice off day (numeric day-of-the-year) from 1970 - present. Add a sentence or two below this exploratory graph describing the overall trend(s) you observe. 
  
### End task
  
  
```{r}
icecover <- read_csv(here("activities", "ntl33_v7.csv"), na = "-999")
weather <- read_csv(here("activities", "ntl20_v6.csv"))
```

```{r}
ggplot(data = icecover, aes(x = lakeid, y = ice_duration)) +
  geom_jitter(width = 0.2, color = "gray50") +
  geom_boxplot(width = 0.3, fill = NA, size = 1, outlier.color = NA)
```


```{r}
mean_duration <- icecover %>% 
  group_by(year4) %>% 
  summarize(mean_duration = mean(ice_duration, na.rm = TRUE))

ggplot(data = mean_duration, aes(x = year4, y = mean_duration)) +
  geom_line()

```


```{r}
weather_summary <- weather %>% 
  filter(month %in% c(12, 1, 2)) %>% 
  group_by(year4) %>% 
  summarize(mean_temp = mean(ave_air_temp_adjusted, na.rm = TRUE))
```

```{r}
ggplot(data = weather_summary, aes(x = year4, y = mean_temp)) +
  geom_line()
```

```{r}
ice_weather_join <- mean_duration %>% 
  full_join(weather_summary)
```

```{r}
ggplot(data = ice_weather_join, aes(x = mean_temp, y = mean_duration)) +
  geom_point()
```
```{r}
thaw_dates <- icecover %>% 
  filter(year4 >= 1970) %>% 
  mutate(ice_off = lubridate::mdy(ice_off)) %>% 
  mutate(day_ice_off = lubridate::yday(ice_off))

ggplot(data = thaw_dates, aes(x = year4, y = day_ice_off)) +
  geom_point()
```
