---
title: "Day 2 Activities"
subtitle: "Bits & bytes, indexing data, import & plot data"
output: 
  html_document:
    css: activity_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
```

## Part 1: Bits & bytes

### Convert the following binary to decimal values: 

- 0010 0111
- 0101 0000
- 0001 1110

### Convert the following decimal values to binary: 

- 243
- 12
- 840

### What does the following (ASCII) spell?

(Use the table from lecture 3) 

110,99,101,97,115

Check your answer using this ascii to text converter: https://convert.town/ascii-to-text

## Part 2. Checking data types

### Setup

- Create a new repo on GitHub called `eds221-day2-activities`
- Clone to create a version controlled R Project 
- Create subfolders called `docs`, `data`, and `figs`
- Create an R Markdown document, save in the `docs` subfolder as `r_data_types.Rmd`

### Create some data, check the classes, index!

#### Vectors, lists & data frames

In your R Markdown document: 

1. Create a vector called `vec_1`containing the following: 

```{r, eval = FALSE}
2, 5, 9, 10, 8, 12, 1, 0
```

```{r, include = FALSE}
vec_1 <- c(2, 5, 9, 10, 8, 12, 1, 0)

vec_1_e5_up <- vec_1[5:7]

vec_1_char <- as.character(vec_1)
vec_1_char

vec_1_logical <- as.logical(vec_1)
vec_1_logical
```

Check the following for that vector: 

- What is the class of the vector? `class()`
- What type of variable does it store? `typeof()`
- Access the 3rd element and store as `vec_1_e3`
- Access the 1st element and store as `vec_1_e1`
- Access the 5th through 7th elements and store as `vec_1_e5to7`
- Reassign `vec_1` as a character using `as.character`, stored as `vec_1_char`. What does the output look like? 

2. Create a vector called `vec_2`

`vec_2` should contained *named* elements, where `town = "Santa Barbara`, `location = "Rincon"`, `swell = "south"`

```{r}
vec_2 <- c(town = "Santa Barbara", location = "Rincon", swell = "south")

class(vec_2)
typeof(vec_2)

vec_2[2]

vec_2["location"]


```

- Take a look at what you've made
- What is the class of vector elements? `class()`
- What is the length of vec_2?
- Access the 2nd element by name and store as `vec_2_e2`

3. Create a data frame in R, index

Write code to create a data frame called `df_1` that looks like this: 
```{r, echo = FALSE}
df_1 <- data.frame(region = c("A", "B", "A", "D"),
                   species = c("otter", "great white", "sea lion", "gray whale"),
                   count = c(12, 2, 36, 6))

df_1

max_count <- max(df_1$count)
```

- Return the class of the entire data frame
- Return the class of the species column
- Find the maximum value of the `count()` column, store as `max_count`


## Part 3. Wild data

#### Set-up

- Visit the EDI site to learn about Mack Creek salamander & cutthroat trout data you'll be using here: [data package](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-and.4027.14)

- Download the *first CSV* listed (`AS00601.csv`), and take a look at it (outside of R is fine as a first step)

- Explore the metadata (see View Full Metadata in the [Resources section](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-and.4027.14) of the data website)

- What does each column contain? What are the units of each? What is the study overall about? 

- Create a new R Markdown document and save it in your `docs` folder. Attach the `tidyverse`, `here` and `janitor` packages in the setup chunk

- Save the `AS00601.csv` in your `data` folder of your project

#### Read in the data

- Read in the data using `read_csv()` with `here()`, store as `mack_verts`
```{r, include = FALSE}
mack_verts <- read_csv(here("activities","AS00601.csv"))
```

- Look at what you've read in

#### A bit of wrangling & exploring

- Update the names to lower_snake_case

```{r, include = FALSE}
mack_verts <- clean_names(mack_verts)

# Or
```

- In a new code chunk, practice accessing individual pieces of the data frame (there is no real functionality to this, but just to reinforce stuff we learned in our interactive session): 

  - Store the 5th value in column "WEIGHT" as `mc_wt_5`. Check by looking at your data frame to confirm.
  - Store the 8th - 20th value in the "LENGTH1" column as `mc_length_8_20`. Check by looking at your data frame to confirm. 
  - Store everything in column SAMPLEDATE as a vector called `mc_dates`
    
```{r, include = FALSE}
mc_wt_5 <- mack_verts$weight[5]

mc_length_8_20 <- mack_verts$length1[8:20]

mc_dates <- mack_verts$sampledate
```

#### Make a salamander subset

- Create a subset that only contains observations for Pacific Giant Salamanders (species *Dicamptodon tenebrosus*, stored in `species` as `DITE`). Store the subset as `mc_salamanders`. *Hint:* see `dplyr::filter()`!

```{r, include = FALSE}
mc_salamanders <- mack_verts %>% 
  filter(species == "DITE")
```

#### Make a scatterplot

- Create a scatterplot of length1 (snout-vent length in millimeters) versus weight (grams) for all salamanders in the dataset. Update axis labels, title, subtitle, and add a caption with the data source. Customize point color and size, possibly opacity, and theme. 

```{r, include = FALSE}
salamander_plot <- ggplot(data = mc_salamanders, aes(x = length1, y = weight)) + 
  geom_point(alpha = 0.5, color = "purple") +
  theme_minimal()

salamander_plot
```

- Export your scatterplot as `salamander_size.png` to your figs folder.

```{r, include = FALSE}
ggsave(here::here("activities", "salamander_size.png"), plot = salamander_plot)
```

#### Make a cutthroat plot

- Similar to above, make a subset called `mc_trout` that only contains observations for cutthroat trout
- Create a scatterplot of length1 by weight for all trout in the dataset
- Customize so that the point color depends on *reach* 
- Customize your color scheme (e.g. `scale_color_manual()`)
- Facet your plot by creek reach (`facet_wrap(~...)`)
- Update graph axis labels and title
- Export your graph as `cutthroat_size.png` to figs

```{r, include = FALSE}
mc_trout <- mack_verts %>% filter(species == "ONCL")

ggplot(data = mc_trout, aes(x = length1, y = weight)) +
  geom_point(aes(color = reach)) +
  scale_color_manual(values = c("red", "orange", "purple")) +
  facet_wrap(~reach) +
  theme_minimal()

ggsave(here("activities", "cutthroat_size.png"))
```

### Stage, commit, pull, push

- Make sure your changes are safely stored by pushing to GitHub
- Close your project completely
- Reopen your project
- Reopen your .Rmd for the salamander & cutthroat work
- Knit. Does it work? Done. 

## End Day 2 activities
