---
title: "Day 1 Activities"
subtitle: "Project organization, data import, file paths"
output: 
  html_document:
    css: activity_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(downloadthis)
library(here)
```

## Setup

### Get the data
 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
stl_blood_lead <- read_csv(here::here("activities","stl_blood_lead.csv"))

stl_blood_lead %>%
  download_this(
    output_name = "stl_blood_lead",
    output_extension = ".csv",
    button_label = "Click to download CSV",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```

Read more about the data [here](https://github.com/slu-soc1120/STL_HEALTH_Lead)
  
### Create your R Project

- Create a new repository on GitHub named `stl_lead_yourname` (mine would be `stl_lead_allison`)
- Clone to create version-controlled R Project
- Add three subfolders to your R project: `data`, `docs` and `figs`
- Copy the dataset you download above into the `data` folder of your project
- Create and save a new .Rmd as `stl_lead_inequity.Rmd` in the `docs` folder

### Read in & explore the data

- Read in the `stl_blood_lead.csv` data as `stl_lead` in your R Markdown
- Do some basic exploration of the dataset, and `janitor::clean_names()` if useful
- In a new code chunk, from `stl_lead` create a new data frame called `stl_lead_prop` that has one additional column called `prop_white` that returns the percent of each census tract identifying as white (variable `white` in the dataset divided by variable `totalPop`, times 100). You may need to do some Googling. Hint: `dplyr::mutate(new_col = col_a / col_b)` will create a new column `new_col` that contains the value of `col_a / col_b`

### Create a scatterplot

- In a new code chunk, create a scatterplot graph of the percentage of children in each census tract with elevated blood lead levels (`pctElevated`) versus the percent of each census tract identifying as white. 
- Customize by updating several aesthetics (e.g. size, opacity (see `alpha = `), color, etc.)
- Store the scatterplot as `stl_lead_plot`
- Have the scatterplot returned in the knitted html - customize the size that it appears when knitted
- Also save a .png of the scatterplot to `figs`, with dimensions of (6 x 5) (width x height)
- In text above or below the scatterplot, write 1 - 2 sentences describing the overall trend that you observe from your graph

### Create a histogram

- Create a histogram of only the `pctElevated` column in the data frame
- Customize the fill, color, and size aesthetics - test some stuff! Feel free to make it awful. 
- Once you've played around with customization, export the histogram as a .jpg to the `figs` folder
- Make sure the histogram also shows up in your knitted html (customize the size)

### Knit

- Knit your document
- Stage, commit, pull then push changes

### Share & test

- Share the link to your public GitHub repo with a neighbor
- Fork the repo your neighbor shared, then clone to get set up locally in RStudio
- Navigate to their .Rmd
- Knit - does everything work? Cool, your neighbor made a reproducible project with file paths you can use too! 

## End activity
