---
title: 'Flex session: basics of reading in data with `readr`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

The goal of this flex session is to read in several different common data types in R using functions in the `readr` package. We'll also see what the analogous Python functions look like for comparison. 

### Wait, what's a flat file? 

We will learn/refresh how to import data from: 

- Comma separated value (.csv) files
- Excel (.xls, .xlsx) files
- Text (.txt) files
- Shapefiles (.shp)
- Data from a URL

## 1. Read in delimited data

The general function in the `{readr}` package (in R, part of the tidyverse) to read in delimited flat files (like comma- and tab-separated value files) is `read_delim()`. The user specifies the delimiter as the second argument in the function. The functions `read_csv()` and  `read_tsv()` are "special cases of the general `read_delim()`" (from R documentation). 

First, let's look at examples for those three functions: 

### `readr::read_delim()`

Read in the `penguins.csv` file using `read_delim()`:

```{r}
penguins_delim <- read_delim("penguins.csv", delim = ",")
```
### `readr::read_csv()`

Since the delimiter is a comma, we can use `read_csv()` directly: 

```{r}
penguins_csv <- read_csv(here("flex_sessions", "penguins.csv"))
```

### Useful options

Check `?read_delim()` and `?read_csv()` to see options for reading in data. Some of my favorites are: 

- `skip = `: how many lines to skip when reading in (default is 0)
- `na = c()`: what values should be converted to `NA` (default is "", "NA")

## 2. Data from an Excel file

Read in data from an Excel file (.xls or .xlsx) using the `readxl` package (in R). Install the package using `install.packages("readxl")`.

Use `read_excel()` to read in data from an excel worksheet (works for both .xls and .xlsx). The default is to read in the *first* worksheet. 

Examples [TODO]

- read in first worksheet
- read in non-first worksheet
- read in & concatenate multiple worksheets

## 3. Get some spatial data, make a map!

