---
title: "Day 2 Interactive Session Materials"
subtitle: "Data types & indexing"
output: 
  html_document:
    css: comp_session_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reticulate)
library(palmerpenguins)
```

# Part 1: Setup

- Create a repo on GitHub called `eds221-day2-comp`
- Clone to make a version-controlled R Project
- Create a new R Markdown, saved in the root as `r-data-types`
- Create a new Jupyter Notebook, saved in the root as `py-data-types`

# Part 2: Making & indexing data in R

## Vectors!

### Making vectors

#### A character vector

```{r}
dogs <- c("teddy", "khora", "waffle", "banjo")

typeof(dogs)
class(dogs)
```

#### A numeric vector

```{r}
weights <- c(50, 55, 25, 35)

typeof(weights) # Hmmm what is different about this and the line below?
class(weights)
```

#### An integer vector

```{r}
dog_age <- c(5L, 6L, 1L, 7L)

typeof(dog_age)
class(dog_age)

# Check with a logical: 
is.numeric(dog_age)
```

#### What if we combine classes? 

There is a hierarchy of classes. The broadest of all in a vector wins (if there are characters, then character will be the class of the entire vector).

```{r}
dog_info <- c("teddy", 50, 5L)
dog_info

typeof(dog_info)
class(dog_info)

is.character(dog_info)
```

#### Named elements

```{r}
dog_food <- c(teddy = "purina", khora = "alpo", waffle = "fancy feast", banjo = "blue diamond")
dog_food

class(dog_food)
typeof(dog_food)
```

### Accessing bits of vectors

Use `[]` with the position or name to access elements of a vector. 

```{r}
dog_food[2]

dog_food["khora"]
```

Or we can specify a range of values within a vector using `[:]`. The first element in **R vectors is assigned element = 1.** This is an important distinction. In Python, the first element is assigned 0 (zero-index). 

```{r}
# Create a vector of car colors observed
cars <- c("red", "orange", "white", "blue", "green", "silver", "black")

# Access just the 5th element
cars[5]

# Access elements 2 through 4
cars[2:4]
```

### A warm-up for for loops:
```{r}
i <- 4
cars[i]

i <- seq(1:3)
cars[i]
```

### And we can update elements of a vector directly (mutable):
```{r}
cars[3] <- "BURRITOS!"
cars
```


## Matrices!

### Creating matrices 

(...we did some of this in EDS 212 too!)

```{r}
fish_size <- matrix(c(0.8, 1.2, 0.4, 0.9), ncol = 2, nrow = 2, byrow = FALSE)

fish_size

typeof(fish_size) # Returns the class of values
class(fish_size) # Returns matrix / array
```

What happens if we try to combine multiple data types into a matrix?
```{r}
dog_walk <- matrix(c("teddy", 5, "khora", 10), ncol = 2, nrow = 2, byrow = FALSE)

dog_walk
class(dog_walk)
typeof(dog_walk)

# Hmmmmmm once again back to the broadest category of data type in the hierarchy
```

### Accessing pieces of matrices

Index using `[row, column]`.

```{r}
whale_travel <- matrix(data = c(31.8, 1348, 46.9, 1587), nrow = 2, ncol = 2, byrow = TRUE)

# Take a look
whale_travel

# Access the value 1348
whale_travel[1,2] # Row 1, column 2

# Access the value 46.9
whale_travel[2,1]
```

If you leave any element blank (keeping the comma), it will return all values from the other element. For example, to get everything in row 2:
```{r}
whale_travel[2,]
```

Or, to access everything in column 1: 

```{r}
whale_travel[, 1]
```

What happens if I only give a matrix one element? That's the position in the matrix *as if populated by column.* Check out a few:
```{r}
whale_travel[3]
```

## Lists

```{r}
urchins <- list("blue", c(1, 2, 3), c("a cat", "a dog"), 5L)

urchins
```
### Accessing pieces of a list

Important: a single [] returns a list. [[]] returns the item STORED in the list. 
```{r}
urchins[[2]]

# Compare that to: 
urchins[2]
```

### Naming list items? Sure thing! 

```{r}
tacos <- list(topping = c("onion", "cilantro", "guacamole"), filling = c("beans", "meat", "veggie"), price = c(6.75, 8.25, 9.50))

# The whole thing
tacos

# Just get one piece of it: 
tacos[[2]]

#...or, the same thing:
tacos$filling

```

## Data frames 

A data frame is a list containing vectors of the same length, where each column is a variable stored in a vector. Let's make one: 

```{r}
fruit <- data.frame(type = c("apple", "banana", "peach"), 
                    mass = c(130, 195, 150))

# Look at it
fruit

# Check the class
class(fruit)
```

### Access elements from a data frame

Use [row#, col#], or name the column (then element number).
```{r}
fruit[1,2]

fruit[3,1]
```


```{r}
fruit[2,1] <- "pineapple"
fruit
```

# Part 3: Making & indexing data in Python

```{python}
import numpy as np
import pandas as pd
```

## Vectors and matrices in Python

```{python}
teddy = [1,2,8]
teddy_vec = np.array(teddy)

teddy_vec
type(teddy_vec)
```

A **list** is mutable - you can change it directly! 
```{python}
teddy[1] = 1000

# See that element 1 is updated directly! 
teddy
```

A **tuple** is immutable - you'll get yelled at if you try to change it! 
```{python}
khora = (1, 5, 12)
type(khora)

# khora[1] = 16 # Nope. 
```

A more involved list (note: you can also use list() to create lists in python).
```{python}
waffle = [["cat", "dog", "penguin"], 2, "a burrito", [1,2,5]]

waffle

# Access an element from the list waffle:
waffle[0] # Default just returns that piece (not as a list)
```

We can reassign pieces of a list: 
```{python}
waffle[1] = "AN EXTRAVAGANZA"

waffle
```

## Make a pandas DataFrame in python

### First, a dictionary example:

```{python}
fox = {'sound': ["screech", "squeal", "bark"], 'age': [2, 6, 10]}

fox['sound']

fox['age']
```



```{python}
cows = {'name': ["moo", "spots", "happy"], 'location': ["pasture", "prairie", "barn"], 'height': [5.7, 5.4, 4.9]}

cows_df = pd.DataFrame(cows)

# Take a look
cows_df

# Get a column
cows_df['name']

# Get an element using df.at[]
cows_df.at[1, 'name']
```

### Side-by-side: R data frame & Pandas DataFrame

```{r}
home_sales <- data.frame(
  state = c("CA", "NV", "OR"),
  sales = c(38000, 4670, 2750)
)

home_sales
```

```{python}
home_sales = {'state': ["CA", "NV", "OR"], 'sales': [38000, 4670, 2750]}

home_sales = pd.DataFrame(home_sales)

home_sales
```


## End (for now...)
