---
title: "Day 3 Interactive Session Materials"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reticulate)
```

## Boolean operator review

## Conditionals

### A basic `if` statement: 

In R: 

```{r}
burrito <- 2.4 # Assign an object value

# Write a short 'if' statement:
if (burrito > 2) {
  print("I love burritos!")
}
```
Try changing the value of burritos to 1.5. What is returned when the `if` statement condition isn't met? 

This is important: for a singular "if" statement, there is no "else" option. So nothing happens if the condition isn't met.

In Python: 

```{python}
burrito = 2.4

if burrito > 2:
  print("I love burritos!")
  
```

#### An example with strings:

```{r}
phrase <- "I love burritos"

if (str_detect(phrase, "love")) {
  print("yay burritos!")
}

```



### A basic `if-else` statement

In R: 

```{r}
pika <- 89.1

if (pika > 60) {
  print("mega pika")
} else
  print("normal pika")
```
In Python: 
```{python}
pika = 89.1

if pika > 60:
  print("mega pika")
else:
  print("normal pika")
```


#### An example with strings: 

```{r}
food <- "I love enchiladas!"

if (str_detect(food, "burritos")) {
  print("yay burritos!")
} else
  print("what about burritos?")

```


### More options: `if-else if-else` statements

In R: 

```{r}
marmot <- 2.8

if (marmot < 0.5) {
  print("a small marmot!")
} else if (marmot >= 0.5 & marmot < 3) {
  print("a medium marmot!")
} else 
  print("a large marmot!")

```

```{python}
marmot = 2.8

if marmot < 0.5:
  print("a small marmot!")
elif 0.5 <= marmot < 3: 
  print("a medium marmot!")
else: 
  print("a large marmot!")
```

### `switch` statements

In R: 

```{r}
species = "mouse"

switch(species,
       "cat" = print("Meow"),
       "dog" = print("Woof!"),
       "mouse" = print("Squeak"))
```



## For loops

### Basic for loop

In R: 

```{r}
dog_names <- c("Teddy", "Khora", "Banjo", "Waffle")

for (pupster in dog_names) {
  print(paste("My dog's name is", pupster))
}

# Or similarly:
for (pupster in dog_names) {
  cat("My dog's name is", pupster, sep = " The Amazing ")
}
```

In Python: 

```{python}
dog_names = ["Teddy", "Khora", "Banjo", "Waffle"]

for i in dog_names: 
  print("My dog's name is " + i)
```

### Another basic for loop example: 

In R: 

```{r}
mass <- seq(from = 0, to = 12, by = 0.5)

for (i in mass) {
  new_val = i + 2
  print(new_val)
}
```

Or, using `seq_along()`:

```{r}
mass <- seq(from = 0, to = 12, by = 0.5)

for (i in seq_along(mass)) {
  new_val = mass[i] + 2
  print(new_val)
}
```

Let's try another one: 

For each element in a vector, find the sum of that value plus the next value in the sequence: 

```{r}
tree_height <- c(1,2,6,10)

for (i in seq_along(tree_height)) {
  val = tree_height[i] + tree_height[i + 1]
  print(val)
}
```

## For loops with conditional statements

A basic conditional for loop in R:

```{r}
animal <- c("cat", "dog", "dog", "zebra", "dog")

for (i in seq_along(animal)) {
  if (animal[i] == "dog") {
    print("I love dogs!")
  } else
    print("These are other animals!")
}

```

Or, for a numerical example: 

```{r}
species <- c("dog", "elephant", "goat", "dog", "dog", "elephant")

age_human <- c(3, 8, 4, 6, 12, 18)

# Convert ages to "animal years" using the following:
# 1 human year = 7 in dog years
# 1 human year = 0.88 in elephant years
# 1 human year = 4.7 in goat years

for (i in seq_along(species)) {
  if (species[i] == "dog") {
    animal_age <- age_human[i] * 7
  } else if (species[i] == "elephant") {
    animal_age <- age_human[i] * 0.88
  } else if (species[i] == "goat") {
    animal_age <- age_human[i] * 4.7
  }
  print(animal_age)
}



```

## Storing outputs of a for loop

To store outputs of a for loop, we'll create an empty vector, then populate it with the for loop elements as they're created. It's important to do this - it will make your loops quicker, which is critical once you start working with big data! 

```{r}
# Create the empty vector animal_ages:
animal_ages <- vector(mode = "numeric", length = length(species))

# Vectors with species and human age: 
species <- c("dog", "elephant", "goat", "dog", "dog", "elephant")

age_human <- c(3, 8, 4, 6, 12, 18)

# Same loop as above, with additional piece added
# To populate our empty vector

for (i in seq_along(species)) {
  if (species[i] == "dog") {
    animal_age <- age_human[i] * 7
  } else if (species[i] == "elephant") {
    animal_age <- age_human[i] * 0.88
  } else if (species[i] == "goat") {
    animal_age <- age_human[i] * 4.7
  }
  animal_ages[i] <- animal_age # Populate our empty vector
}

```

Another example of storing an output: 

```{r}
tigers <- c(29, 34, 82)
lions <- c(2, 18, 6)

big_cats <- vector(mode = "numeric", length = length(tigers))

for (i in seq_along(tigers)) {
  total_cats <- tigers[i] + lions[i]
  big_cats[i] <- total_cats
}

```

Don't make your life harder for no reason. What's the easiest way to fine the `big_cats` values as calculated above? 





