---
title: "Day 5 Interactive Session Materials"
subtitle: "Functions continued: testing, sourcing, documentation, sharing"
output: 
  html_document:
    css: comp_session_theme.css
    toc: true
    toc_float: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(testthat)
library(reticulate)
library(palmerpenguins)

source("food_cart_functions.R")
```

## Setup

-   Fork then clone [this repo](https://github.com/allisonhorst/eds221-day5-comp) to create a version-controlled R Project for Day 5

## Part 0: Warm up - For loops revisited (Leslie Matrix)

Open the `leslie_projection.Rmd` file in docs/. All code shown below for Part 0 is in that file. You should be able to run it out-of-the-box. In EDS 212, we learned about Leslie matrices for projecting populations based on mortality and reproduction for different organism life stages. Let's return to one example we checked out. 

A population of fish we're interested in has four life stages: eggs (E), fry (F), juvenile (J), breeding adult (A). You are told that annually:

- Each breeding adult will produce, on average, 200 eggs
- 5% of eggs will survive to become fry
- 10% of fry will survive to become juveniles
- 16% of juveniles will survive to become adults
- Adult survival rate year to year is 90%

### Create the matrix

```{r}
# Make the Leslie Matrix:
fish_leslie <- matrix(c(0, 0, 0, 200, 0.05, 0, 0, 0, 0, 0.10, 0, 0, 0, 0, 0.16, 0.9), nrow = 4, ncol = 4, byrow = TRUE)

# Check it out: 
fish_leslie
```

Recall, we are able to use the dot product (`%*%`) to estimate the population at the next year (we'll draw this on the board to remember what it looks like). 

The initial population structure is given by: 
```{r}
# Initial population structure (Roe, Fry, Juvenile, Adult): 
fish_year0 <- c(5000, 8000, 600, 400)


# ------ WHAT HAPPENS HERE? ------ #
time_yr <- seq(from = 0, to = 8, by = 1)
proj_year <- matrix(ncol = length(fish_year0), nrow = length(time_yr))
proj_year[1, ] <- fish_year0


for (i in 2:length(time_yr)) {
  proj_year[i,] <- fish_leslie %*% proj_year[i-1,]
}
# -------------------------------- #

# The rest is wrangling & visualization:
colnames(proj_year) <- c("eggs", "fry", "juvenile", "adult")
proj_df <- data.frame(time_yr, proj_year) %>% 
  pivot_longer(cols = -time_yr, names_to = "lifestage", values_to = "stage_population")

ggplot(data = proj_df, aes(x = time_yr, y = stage_population)) +
  geom_line(aes(color = lifestage)) +
  scale_y_log10()

```

## Part 1: Testing functions with `{testthat}`

-   Create a new R Markdown file in your existing project (in docs/) saved as `fun_testing.Rmd`
-   Install the `testthat` package (`install.packages("testthat")`)
-   Install the `devtools` package (`install.packages("devtools")`)
-   Install the `remotes` packages (`install.packages("remotes")`)

As we develop algorithms, we'll change our code. We want a way to automatically check our work to make sure it's behaving as expected. The [`testthat`](https://testthat.r-lib.org/index.html) package "tries to make testing as fun as possible."

**Unit test:** A unit test is an automated check of a piece ("unit") of your code

Let's consider an example. We're writing a function to find the mean value of each column, then return the lowest and highest mean calculated (in that order). We would expect:

-   The outcome to be a numeric vector of length 2
-   Where the first value in the vector is smaller than the second value in that vector

### Write the function

Let's write the function first, then create some accompanying tests to make sure they're working.

```{r}
mean_range <- function(df) {
  col_means <- apply(X = df, MARGIN = 2, FUN = mean, na.rm = TRUE) # Returns column means as a vector
  col_mean_max <- max(col_means) # Looks for the maximum value in the vector
  col_mean_min <- min(col_means) # Looks for the minimum value in the vector
  return(c(col_mean_min, col_mean_max)) # Prints the vector with minimum & maximum
}

# Try it out:
mean_range(df = mtcars)
```

OK great. But we're doing some work on this code, and we don't want to have to try a bunch of different things manually each time we change it to see how it breaks. Instead, let's write some **automated tests** for this function that help us avoid that.

### Create some tests with `testthat`

The `testthat` package contains a whole bunch of helper functions built for testing. Take a look at [options in the package documentation](https://testthat.r-lib.org/reference/index.html).

Here are some examples of the types of built-in functions for testing:

-   `expect_length()`: does the function return a vector of the expected length?
-   `expect_equal()`: does the function return an expected value?
-   `expect_true()`: does the code return `TRUE` or `FALSE`?

For example, let's use `expect_length()` to test our function above, which should return a vector of length 2.

```{r}
expect_length(mean_range(mtcars), 2)
```

Nothing is returned - that's passing. You'll get an error message if the test fails. What if we say the expected length of the vector returned is 3?

```{r, error = TRUE}
expect_length(mean_range(mtcars), 3)
```

Let's do another one!

For this function to be working correctly, the output of `mean_range(mtcars)` should have two values and we expect the **first element to be smaller than the second element.**

Our test might look something like:

```{r}
expect_true(mean_range(mtcars)[1] < mean_range(mtcars)[2])
```

That passes! What happens if you change the alligator mouth direction?

```{r, error = TRUE}
expect_true(mean_range(mtcars)[1] > mean_range(mtcars)[2])

```

That test fails. This is just a little flavor of what unit tests do.

## Part 2: Sourcing functions

When you're writing functions (or working with someone else's functions), often they're not stored in the actual script or notebook you're working in. If that's the case, you can still use them by **sourcing** a function.

Let's make a function that's stored in a script, save it in our project working directory, and use it in an R Markdown document.

-   Create a new R script (File \> New File \> R Script)
-   Save the script in your project

In our new R script, let's write a function that, when given inputs of the user's favorite food and animal, returns the name of their future food cart.

```{r, eval = FALSE}
name_cart <- function(food, animal) {
  print(paste("Mc", stringr::str_to_title(animal), "'s ", stringr::str_to_title(food), "Mart", sep = ""))
}

# Try it out: 
name_cart(food = "burrito", animal = "pika")
```

Save the script with a descriptive name, e.g. `food_cart_functions.R` (this could, in theory, contain multiple functions). Atop our .Rmd (in the setup chunk), we'll **source** the script so we have access to its functions. Mine will look like this:

`source("food_cart_functions.R")`

**Note:** you may have your scripts in a subfolder (e.g. R/ or something) - you can use `here::here()` within the `source()` function to point to the right place within your R Project as needed.

Now, add a code chunk in your R Markdown document that uses the `name_cart()` function you've created. It works over here, too!

```{r}
name_cart(food = "strawberry", animal = "marmot")
```

This gives us a valuable tool for organizing things within a project. Often, you'll have a separate folder called `R` where you have your scripts / functions, and those are sourced as needed for use in your notebook or by another script you're working on.

### A real sourcing example: cork oak growth

**Citation:** Mariola Sánchez-González, Margarida Tomé, Gregorio Montero. Modelling height and diameter growth of dominant cork oak trees in Spain. Annals of Forest Science, Springer Nature (since 2011)/EDP Science (until 2010), 2005, 62 (7), pp.633-643.

In the `src/` folder there is a script `cork_oak_growth.R` that contains one function `height_t2` that estimates the height of a cork oak tree at some age (time 2), given a prior height of the tree, the age when that height was recorded (years), and the age you're trying to estimate the new height for (these ages are t1 and t2, respectively, in the model). 

Create a new .Rmd. Source the script by running `source(here::here("path", "to", "script.R"))` in your setup chunk. Ensure that you can use the function in your .Rmd whenever you want to. 

## Part 3: Function documentation

Documentation is a really important part of data science. Remember: "past you doesn't respond to emails" (Wickham). I'll also add that present you will forget things immediately, and future you is a grump. So make sure, for your sake and the sake of your collaborators, you **document** your functions (along with any useful annotation throughout your work) to make them easier for you to use later on.

It's important to note here that documentation can mean different things.

Informally, this can mean adding relevant comments to your script so that you know:

-   Arguments expected (name and units)
-   Variable types expected (and output)
-   A summary of what the function does
-   An example of how it works
-   Any citations / relevant information

Formally, in R, the `{roxygen2}` package can help guide our documentation.

Let's try creating some documentation for a little function. Write a function to calculate the maximum sustained harvest ($H$), given a fishery carrying capacity ($K$) and intrinsic growth rate, $r$:

$$H = \frac{Kr}{4}$$

```{r}
# Write the functio
max_sustainable_harvest <- function (K, r) {
  harvest <- (K * r) / 4
  return(harvest)
}

# Try it out: 
max_sustainable_harvest(K = 36000, r = 0.31)
```

Now, how can we document this function?

We'll use what's called a *Roxygen Skeleton* - a skeleton set of documentation that prompts us to input important information about our function. **Note:** this exercise is just so you get a feel for what a roxygen skeleton looks like.

To add a Roxygen Skeleton:

-   Click anywhere within your function
-   In RStudio, go up to 'Code' in the top menu, and click 'Insert Roxygen Skeleton'
-   Immediately above your function, you should see some auto-added text that looks like this (your parameters will already be added):

```{r, eval = FALSE}

#' Title
#'
#' @param  
#' @param  
#'
#' @return
#' @export
#'
#' @examples

```

-   Update the fields with descriptions and types

    -   `@param`: information about the parameters (arguments) in your function
    -   `@return`: information about what the function *returns* to the user
    -   `@export`: we'll return to this, but this lets an R package know that this is a function available for the user to use (not an internal function)
    -   `@examples`: add helpful examples for the user

```{r, eval = FALSE}
#' Maximum sustainable harvest
#' 
#' A little function to return the maximum sustainable harvest given the carrying capacity and growth rate.
#'
#' @param K A number indicating the carrying capacity of the fishery (in individuals)  
#' @param r A number indicating the intrinsic growth rate of the fishery (in 1 / yr)
#'
#' @return A number indicating the maximum annual sustainable harvest for the fishery
#' @export
#'
#' @examples
#' max_sustainable_yield(K = 3.8e5, r = 0.62)
```

So overall, our function code chunk might look like this:

```{r}
#' Maximum sustainable harvest
#' 
#' A little function to return the maximum sustainable harvest given the carrying capacity and growth rate.
#'
#' @param K A number indicating the carrying capacity of the fishery (in individuals)  
#' @param r A number indicating the intrinsic growth rate of the fishery (in 1 / yr)
#'
#' @return A number indicating the calculated maximum annual sustainable harvest for the fishery
#' @export
#'
#' @examples
#' max_sustainable_harvest(K = 3.8e5, r = 0.62)

max_sustainable_harvest <- function (K, r) {
  harvest <- (K * r) / 4
  return(harvest)
}

# Try it out: 
max_sustainable_harvest(K = 36000, r = 0.31)
```

Now, this might seem weird because it's not obvious how that odd formatting of the Roxygen Skeleton actually helps. Why not just make them normal comments, and it words that a human can easily read?

The answer is: because this structure and syntax will make it possible for us to store our documentation as part of an R package containing our function. And that's what we want to do next...

## Part 4: Making an R package

Let's put a bunch of what we've been working on together into an R package that you can use, *and* share with others so they can use your amazing function!

Step-by-step:

1.  In RStudio, new R Project \> New directory \> R package

2.  Name your R package something awesome, like your favorite food and favorite animal (e.g. tacopika)

3.  Save your new package project in your EDS 221 folder, and open your project in a new session

4.  Hey, you made an R package!

    -   Some placeholder example stuff exists in it already
    -   Including a single function called `hello()` that prints the worlds "Hello, world!"
    -   The function is saved in a single .R script, with the same name, within the `R` folder

5.  Wait you made an R package? Yup try it out.

    -   Go up to the Build tab, and click 'Install and restart'
    -   You should see that your package is attached (see `library(your_package_name)` in the Console)
    -   Now you can use any of the exported functions - which so far is only `hello()`. Try it!

6.  OK but let's *really* make it your own...

  -   Create a new R script in your project

  -   Create a function within the script of your choosing (you'll share this with a classmate). Some requirements:

      -   It must have at least 3 parameters
      -   Of those 3 parameters, at least one must be a character, and one numeric

  -   Save the script **in the R subfolder with the same name as your function.** For example, if your function is called `jurassic_park()`, then the script should be saved in the R folder as **jurassic_park.R**.

  -   Click 'Install and restart' in the Build tab again

  -   Try your new function!

7.  Documenting your function

So that's cool that your new function exists within the package and you can use it. But now try checking the documentation on your function using `?function_name`. What shows up? Zilch, right?

We want to add documentation for our function that we can check it out whenever we use the function, and so that other users will have critical information about what it does and how it works.

-   In the script with your function, click anywhere within your function
-   Add a roxygen skeleton (Code \> Insert Roxygen Skeleton)
-   Update your documentation with at least a new title and parameter descriptions. Save.
-   Run `devtools::document()` (note: if you see 'Warning: Namespace not generated by Roxygen2', you should delete the existing NAMESPACE file, then run `devtools::document()` again to resolve)
-   Something new is created - check the `man/` folder and see that an .Rd (R documentation) file has been added that is associated with your function name, and that contains your updated documentation
-   Press Install and restart in the Build tab
-   Check for your function documentation - does it exist now? Cool!

Things to remember:

-   Run `devtools::document()` any time you update your function or documentation
-   You must use `Install and restart` to see the changes when you test

8.  Sharing your package

**Putting it on GitHub**

Your friends and family will definitely want to use the amazing R package you've just created. How can *they* install and use it?

Collaborators can install your package straight from a GitHub repo. Let's put it up there.

-   In your package project, run `usethis::use_git()`
-   Then run `usethis::use_github()`

This should take you to your new GitHub repo for the package.

**Installing a package from GitHub**

Once your package is in a public repo on GitHub, anyone can install its contents by running:

    `remotes::install_github("username/reponame")`

For example, if my package is in a repo called `tacopika`, then someone could install that package of mine using:

    `remotes::install_github("allisonhorst/tacopika")`

**Note:** you can check all functions in a package by running `lsf.str("package:packagename")` in the console. For example, after installing the package `tacopika` above, running `lsf.str("package:tacopika")` in the Console will return all functions in it. 

## Part 5. Finding, vetting and using other packages

Now that you've made your own package, let's dive into packages that *other* developers have made so that we do more data science and less reinventing the wheel with code to do common useful things.

Let's spend a bit of time learning how to search for, choose, investigate put to use some R & Python Packages.

### Finding packages

-   CRAN Task Views (search by topic): <https://cran.r-project.org/web/views/>

-   Microsoft R Application Network (MRAN) (search by keywords): <https://mran.microsoft.com/packages>

-   Some trusted & awesome package sources:

    -   ROpenSci: <https://ropensci.org/packages/>
    -   RStudio: <https://www.rstudio.com/>
    -   pyOpenSci: <https://www.pyopensci.org/>

-   Other great places to learn about package updates, releases, use:

    -   \#rstats on Twitter (seriously)
    -   [R Views blog](https://rviews.rstudio.com/) by Joseph Rickert
    -   Also...just a lot of googling (and using useful keywords, e.g. "tidyverse")

### Exploring / vetting packages

**Explore packages:**

-   Usually, on their GitHub repo
-   Contributors
-   Activity (recent commits?)
-   Issues (and responses)?
-   How many packages depend on this packages?

**Dig a bit deeper:**

-   Source code
-   pkgdown website
-   relevant citations

### Using and citing packages

**Installation**

We've seen the two main ways you'll install R packages:

-   From CRAN (using `install.packages("packagename")`)
-   From GitHub (using `devtools::install_github("username/packagename")` or `remotes::install_github("username/packagename"`)

**Note:** There are other ways to get R packages. For example, [Bioconductor](https://www.bioconductor.org/) provides "open source software for bioinformatics" and is home to a lot of great software & packages for genomics work. See the Bioconductor [installation page](https://www.bioconductor.org/install/) for information about how to install their packages & tools.

What about Python packages? We haven't dealt with this at all, since *most* of the packages we need are installed with Anaconda. What id we need something different?

Follow along to make sure you can install packages with pip:

-   Open the Terminal

-   Check for pip by running the appropriate commands (make sure you choose the correct one for your OS)

-   Install pip if necessary (follow along with commands [here](https://packaging.python.org/tutorials/installing-packages/#ensure-you-can-run-pip-from-the-command-line))

-   Install Python packages by running the following in the command line, replacing "project_name":

    -   MacOS: `python3 -m pip install "project_name"`
    -   Windows: `py -m pip install "project_name"`

For example, I have pip installed. To install the `seaborn` package, I run: `python3 -m pip install "seaborn"`

**Go ahead and try it:** Use the appropriate command for your OS to install the Python version of the `palmerpenguins` package (also named `palmerpenguins`).

**Use and responsibility**

Anyone can make and publish an R or Python package to GitHub. That means it is *your* responsibility to determine if it is safe to use for your science. Once you've sufficiently explored, vetted and tested a package, the next step is to use it.

**Citing packages**

Open source software is a major contribution to science and deserves to be credited. Often, developers are creating and maintaining packages voluntarily and without pay or support. Please remember to cite software and tools that helps you with your work.

Luckily, packages (at least those on CRAN) come with a built-in citation to make it easy for you to see the citation. Use `citation(packagename)` to return package citation.

For example:

```{r}
citation("palmerpenguins")
```

And an aside: R Markdown now makes it pretty nice to interface between your Zotero libraries and your R Markdown document. In the Visual Editor mode for R Markdown, you should see an \`\@\` symbol. Clicking on it (if you have Zotero) should bring up your libraries.\
\
Select the citation you want to insert. For example, data used in this example are from the `palmerpenguins` R package [@horst2020a].

Click on the citation text for a preview of how it'll show up in the final document!

Learn more about citations in R Markdown: <https://blog.rstudio.com/2020/11/09/rstudio-1-4-preview-citations/>

## END

Extra unit test example (from lecture):

```{r, error=TRUE}
fish_mass <- function(alpha, beta, fish_length) {
  
  if (fish_length <= 0) {
    stop("Fish length must be greater than or equal to 0.")
  }
  
  mass <- alpha * (fish_length ^ beta)
  return(mass)
  
}

# Unit tests

expect_length(fish_mass(alpha = 1, beta = 2, fish_length = 3), 2)
```

Extra Roxygen example:

```{r}
#' Calculate the distance traveled in miles
#'
#' @param rate a number in units of miles per hour
#' @param time a number in units of hours
#'
#' @return a number indicating the distance traveled in miles
#' @export
#'
#' @examples
distance <- function(rate, time) {
  dist <- rate * time
  return(dist)
}
```
