---
title: "Day 10 Interactive Sessions"
subtitle: "Updating our R packages, kable tables, a first git collaboration"
output: 
  html_document:
    css: comp_session_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(lubridate)
library(kableExtra)
library(palmerpenguins)
```

## 0. Setup

- Create new repo on GitHub named `eds221-day10-comp`
- Clone to make a local version-controlled R Project

## 1. Make a ggplot theme

- Add a new R Markdown document, saved in your project root as `my_ggplot_theme.Rmd`
- Attach the `tidyverse` and `palmerpenguins` packages
- Create a plot from the data (whatever type you want)
- Highly customize the `theme()` component (you can make it as bright / awful as you want - this is going to become a ggplot theme you can share with the world, so it's up to you)
- Keep this project OPEN (you'll copy your custom theme in the next step...)

Here's something awful just to remind you of what this can look like:

```{r, eval = FALSE}

ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  theme(title = element_text(size = 16, color = "purple"),
        plot.background = element_rect(fill = "black"),
        panel.background = element_rect(fill = "gray20"),
        axis.text = element_text(color = "yellow"),
        panel.grid.major = element_line(color = "blue"),
        panel.grid.minor = element_line(color = "cyan")
        )

```

## 2. Updating our R package

- In a new session (so your existing project stays open), reopen the R Project for your R package you created last week
- Install and Restart your package (this should also attach your package)
- Remind yourself of what functions exist in the package (one way: in the Packages tab, click on your package name to see a list)
- Create a new R script (.R )
- Copy and paste just the `theme()` component of your customized ggplot graph that you just made into your empty R script. For the example above, that would just be: 

```{r, eval = FALSE}
theme(title = element_text(size = 16, color = "purple"),
        plot.background = element_rect(fill = "black"),
        panel.background = element_rect(fill = "gray20"),
        axis.text = element_text(color = "yellow"),
        panel.grid.major = element_line(color = "blue"),
        panel.grid.minor = element_line(color = "cyan")
        )
```

- Put that theme inside of a function, and assign it a name. For example: 

```{r, eval = FALSE}
theme_eighties <- function() {theme(title = element_text(size = 16, color = "purple"),
      plot.background = element_rect(fill = "black"),
      panel.background = element_rect(fill = "gray20"),
      axis.text = element_text(color = "yellow"),
      panel.grid.major = element_line(color = "blue"),
      panel.grid.minor = element_line(color = "cyan")
)
}
```

- Save your script **using the same name as the function** in the `R` folder (e.g. this one would be `theme_eighties.R`)
- Add a Roxygen skeleton (recall: click within your function > Code > Insert Roxygen skeleton), and update the title (note that there aren't any arguments / params)
- `devtools::document()` to produce the R documentation for your new function
- Install & restart
- Try it out! Back in your *other* project, attach your package (which you just reinstalled), then make a plot that uses your custom theme from your package! 

For example, if mine is called `tacopika`: 

```{r, eval = FALSE}
library(tacopika)

ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  theme_eighties()
```

- Stage, commit, & push changes back to your GitHub repo
- Share your repo information (username/reponame) with your neighbor so they can install your package from GitHub! Recall: `devtools::install_github("username/reponame")`
- Test your neighbors new & improved R package and try out their custom theme! 

## 3.  Making some nice tables in R

We've made a number of ggplot graphs, but we haven't made any tables. Let's learn one way! 

- Back in your `eds221-day10-comp` R Project, create a new R Markdown document
- In the setup chunk, attach the `tidyverse` & `kableExtra` (note: there are a *bunch* of ways to make nice tables in R - see David Keyes' post on [*How to make beautiful tables in R*](https://rfortherestofus.com/2019/11/how-to-make-beautiful-tables-in-r/) for more options)
- Copy the contents of the table below to the clipboard, the use the `datapasta` Add-in to create a tibble stored as `whale_sightings`

```{r, echo = FALSE}

dt <- tribble(
  ~date, ~site, ~spp, ~dist_m, ~behavior,
  "8/12/2014", "channel", "unknown", 400, "breach",
  "8/13/2014","channel", "gray", 200, "spout",
  "8/15/2014","harbor", "gray", 60, "spout",
  "8/16/2014","channel", "humpback", 300, "feeding",
  "8/16/2014","channel", "gray", 150, "feeding"
)

whale_sightings <- dt

dt %>% 
  kable() %>% 
  kable_styling(full_width = FALSE)
```

#### Let's make some nice looking tables

With `kableExtra`: 
```{r}
# Bootstrap theme
dt %>% 
  kable(col.names = c("Date", "Site", "Species", "Distance (m)", "Behavior")) %>% 
  kable_styling(full_width = FALSE, bootstrap_options = "striped")

# Paper theme
dt %>% 
  kable() %>% 
  kable_classic()
```

Check out some other themes and try them out! 
https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html

A bit more customization: 
```{r}
dt %>% 
  kable(col.names = c("Date", "Site", "Species", "Distance (m)", "Behavior")) %>% 
  kable_classic() %>% 
  column_spec(1, bold = TRUE, background = "yellow") %>% 
  column_spec(2, italic = TRUE, background = "orange") %>% 
  add_header_above(c(" " = 1, "One header" = 2, "Another header" = 2)) %>% 
  scroll_box(height = "100px", width = "500px")
```

See also: `DT`, `reactable`, `gt`, and more! 

### 4. Git collaboration: both pushing to main

- Find a partner. Designate who is "Starter" and who is "Collaborator". 

#### Starter: Create a new version controlled R Project, make & push an update

- Create a new repo called eds221-day10-collab (with a ReadMe)
- Clone, create a new version-controlled R Project
- Create a new .Rmd in the root (`git_test.Rmd`), delete everything below the first code chunk
- Add a line of text in your .Rmd like “Hi partner!”
- Stage, commit & push all changes back to your GitHub repo

#### Starter: Add a contributor

- Go back to your eds221-day10-collab repo on GitHub
- Go to Settings > Manage access > (Enter password if requested) > Invite collaborator > Enter Collaborator’s username or email

#### Collaborator: Accept invitation & clone repo, make an update

- Check your email. You should receive the invitation to join the repo. Accept.
- Clone (do NOT fork first) and create your own local R project
- Pull just in case (this should say already up to date)
- Open the git_test.Rmd 
- Add a new line to the .Rmd with a nice note below the line your partner added
- Save the .Rmd, then stage, commit, pull & push 
- Check that the updates show up on GitHub

#### Starter: Pull & add something new to the .Rmd

- PULL to get remote updates locally in RStudio
- Open git_test.Rmd and see updated text from your partner
- Add a new line of text to the .Rmd
- Stage, commit, pull, then push

#### Collaborator: Pull & add something new to the .Rmd

- PULL to get remote updates
- Open git_test.Rmd and see updated text
- Add a new line of text to the .Rmd
- Stage, commit, pull, then push

### Where goes what now on GitHub?

- Both partners, go back to the repo on github for your collaboration
- Go exploring (especially History & Blame)

### Submit a new issue that references specific line(s) in your files

- Still in GitHub, click on the git_test.Rmd file
- Find a line that your partner wrote
- Click the row number to the left of the line (or hold shift to select a range a lines) - the selected lines will turn yellow, and you should see a three dot menu button appear to the left of the code. Click on it, and choose "Reference in new issue"
- Add a title and some text for your issue, and submit
- Once you both have submitted an issue to your shared repo, check your partner's issue, respond and close (resolve) the issue

