---
title: "Day 4 Interactive Session Materials"
subtitle: "Functions - introduction, basics, examples"
output: 
  html_document:
    css: comp_session_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reticulate)
library(palmerpenguins)
```

## Part 1: Learning from existing functions

### Function documentation

Let's check out documentation for a few basic functions in R. This will help us get an idea of the type of features we may want to add to our functions in the future, and to familiarize ourselves with how to read and understand function documentation (which can admittedly be...a lot). 

#### In R...

To see documentation for R functions, run `?functionname`. This should bring up the function documentation in the Help tab in RStudio. For example, try running `?min`. 

What are the different pieces in the documentation? 

- **Description:** gives an overview of what the function does
- **Usage:** shows general usage and default behaviors
- **Arguments:** the arguments...
- **Details:** any other details...
- **Value:** information about the output returned
- **Examples:** selected examples (but also...Google it)


#### In Python...

To see function documentation in Python, run `help(functionname)`. This will bring up a new window with information about the function (press `q` to escape). For example, try running `help(min)` in the Python interpreter. 

## Part 2: Basic pieces of a function

### Components of a function

From [Chapter 19 - Functions](https://r4ds.had.co.nz/functions.html) in R for Data Science: "There are three key steps to creating a new function:

- You need to pick a name for the function. Here I’ve used rescale01 because this function rescales a vector to lie between 0 and 1.

- You list the inputs, or arguments, to the function inside function. Here we have just one argument. If we had more the call would look like function(x, y, z).

- You place the code you have developed in body of the function, a { block that immediately follows function(...)."

## Part 3: Getting started - function examples in R & Python

#### Example 1

Make a function called `double_it` that doubles any value input value. 

In R: 

```{r}
# Create the function: 
double_it <- function (x) {
  2 * x
}

# Try it out! 
double_it(24) 

# And on a vector, still works: 
double_it(c(1, 0.5, -200))
```

In Python: 

```{r}

```

#### Example 2

Write a function that returns the statement "I am ___ years old!", where the blank is entered by the user as argument `age`. 

In R: 
```{r}
my_age <- function(age) {
  print(paste("I am", age, "years old!"))
}

my_age(age = 10)
```

In Python: 

```{python}

```

## Part 4: Functions with conditionals

Similar to for loops, we can have conditionals within functions - this means that *what* the function does with arguments will change based on what those argument values are. 

#### Example

Returning to an example that we saw previously, let's say we want to make a function that returns the age of an animal in animal years, given the species and age in human years. For example: 1 human year = 7 dog years, and 1 human year = 4.7 goat years. 

Write a function with two arguments: `animal` and `age`, which converts the animal's age from human to animal years. 

In R: 

```{r}
animal_age <- function(animal, age) {
  if (animal == "dog") {
    print(age * 7)
  } else if (animal == "goat") {
    print(age * 4.7)
  } 
}

# Try an 8 year old dog.
animal_age(animal = "dog", age = 8)


# Try a cow & see what is returned. What happens? Consider.
animal_age(animal = "cow", age = 12)
```

**What if the input isn't dog or goat?** Remember this question - we'll return to it later on. 

**A question to keep in mind:** it's easy enough to manually code in the animal species and conversion factor since we only have two animals. Would we want to manually write this out if we had 20 species? 50 species? 

#### Example

Let's say we have a table of the dogs' favorite foods. 

```{r, echo = FALSE}
dog_choice <- tribble(
  ~dog_name, ~food,
  "Khora", "everything",
  "Teddy", "salmon",
  "Waffle", "pancakes",
  "Banjo", "chicken"
)

dog_choice
```

Write a function that only requires the dog name as an input, then automatically pulls information from the data frame (stored as `dog_choice`) to return the phrase "My name is _____ and I love eating ____!"

Here I'll introduce a new function: `dplyr::filter()`. This is a fantastic function for creating subsets from our data. We're going to be using it a *LOT* in our data wrangling lessons last week. For now, know that it returns a logical TRUE or FALSE based on whether or not elements in a vector satisfy the conditions. 

Let's talk through what this code is actually doing: 

```{r}
dog_menu <- function(name) {
  my_sub <- dplyr::filter(dog_choice, dog_name == name)
  print(paste("My name is", my_sub$dog_name, "and my favorite food is", my_sub$food))
}

# Try it! What is Waffle's favorite food?
dog_menu(name = "Waffle")

# And Khora? 
dog_menu(name = "Khora")
```
Aha - now we can see that we don't have to write out every value in our function! We can pull elements directly from a table to switch parameters based on user selection / inputs!

## Part 5: Functions meet for loops

Sometimes, we'll want to iterate over different elements of a data structure (e.g. data frame columns), applying a function we write to each of them. We've learned a couple of coder-friendly tools to loop over columns (e.g. `apply`, `purrr::map()`), but let's write one from scratch first. 

#### Example

Write a function that iterates over each column in a data frame (name gets specified), calculating the mean value for each and returning the phrase "The mean value of (column name) is _____." 

Note: the `colnames()` function in base R returns column names. 

```{r}
df_means <- function(df) {
  for (i in 1:ncol(df)) {
    col_mean <- mean(df[[i]])
    column_name <- colnames(df[i])
    print(paste("The mean value of", column_name, "is", col_mean))
  }
}

df_means(df = mtcars)
```

**Question:** Why only the single square brackets to refer to the column names here? Try it! What is returned when we use `colnames(mtcars[2])` versus `colnames(mtcars[[2]])`?

So, will this actually work on other data frames? Let's make one! 

```{r}
the_fish <- tribble(
  ~brown, ~rainbow, ~cutthroat,
  16.2, 12.1, 21.9,
  24.7, 14.6, 18.3,
  12.1, 15.4, 30.2,
  10.4, 11.8, 26.3
)

df_means(df = the_fish)
```
#### Another example

A function containing `switch()`? See https://statisticsglobe.com/switch-function-in-r/

## Part 6: Adding helpful error and warning messages

We now know a bit about writing functions. But we also want them to be as *helpful* as possible - meaning that we need to create useful error messages, documentation and tests. We'll start by adding some `stop()` statements to create error messages. 

First, it's important when designing a function to think about what will make it fail. For example, if a function does an age conversion on a number, what happens if the user inputs a character? What happens if a user only enters one argument value, but the function requires two? Let's take a look at some examples for adding error messages. 

### Adding error messages with conditional checks + `stop()`

Returning to our function to convert animal ages: 

```{r, echo = TRUE, eval = FALSE}
animal_age <- function(animal, age) {
  if (animal == "dog") {
    print(age * 7)
  } else if (animal == "goat") {
    print(age * 4.7)
  }
}
```

We want to add some error messages that are helpful to the user of the function. We'll use the `stop()` function 

```{r}
animal_age_stop <- function(animal, age) {
  
  if (!animal %in% c("dog", "goat")) {
    stop("Oops! Animal must be a dog or a goat.")
  }
  
  if (is.numeric(age) == FALSE) {
    stop("The age must be a number between 0 and 100")
  }
  
  if (age <= 0) {
    stop("Age must be a number greater than zero.")
  }
  
  if (animal == "dog") {
    print(age * 7)
  } else if (animal == "goat") {
    print(age * 4.7)
  }
}
```

### Adding warning messages

Sometimes, you still want code to *run* (i.e., not throw an error message and stop), but there might be something suspicious about the inputs or outputs that you'd want to alert the user about. For example, maybe you'll flag an input for dog age if the person adds a value over 22 years - it won't *not* work, but it'll give a friendly heads up that that's a very old dog, are you sure that's the value you wanted to enter? 

We can add **warning messages** to functions in R using `warning()`. Keep in mind that the function will still run - this is very different than a hard stop from an error message. 

**Example:** Betz' Limit

The full power in wind hitting a turbine is: 

$$P = 0.5\rho Av^3$$

where $P$ is power in Watts (joules/second), $\rho$ is the air density (kg/m^3^), $A$ is the area covered by the turbine blades (square meters), and $v$ is the wind velocity (m/s). 

However, the [Betz Limit](https://energyeducation.ca/encyclopedia/Betz_limit) means that turbines can only collect ~60% of the total wind power, which updates the theoretical "collectable" power (before accounting for inefficiencies, losses, etc.) to: 

$$P = 0.3\rho Av^3$$

Write a function to calculate *maximum collectable* wind power (Watts) by a turbine requiring three inputs: 

- Air density (in kg/m^3^)
- Rotor radius (in meters)
- Wind velocity (in m/s)

Add the following errors and warnings:

- A warning if the windspeed entered is > 130 (m/s)
- A warning if the air density is > 1.225 kg / m^3^ (air density at sea level)
- An error if the rotor radius is < 0 

```{r}
calc_windpower <- function(rho, radius, windspeed) {
  
  if (windspeed > 130) {
    warning("Whoa, that's a high windspeed! Are you sure that's correct?")
  }
  
  if (rho > 1.225) {
    warning("That air density is suspicious.")
  }
  
  if (radius < 0) {
    stop("Rotor radius must be a positive value (meters).")
  }
  
  0.3*rho*pi*(radius^2)*(windspeed^3)
  
}
```

Now test your function with different inputs. Are your messages working? 

## Part 7: Testing functions with `{testthat}`

As we develop algorithms, we'll change our code. We want a way to automatically check our function to make sure it's behaving as expected. The `testthat` function creates a fun way to create unit tests for your code. 

**unit test:** TODO definition

Let's say, for example, we're writing a function to find the mean value of each column, then return the lowest and highest mean calculated (in that order). We would expect: 

- The outcome to be a numeric vector of length 2
- Where the first value in the vector is smaller than the second value in that vector



